Space transformation in Multidimensional IRT (Lemmas)

### 2022-08-11 notebooks/Space_transformation_theorem.qmd
Direction cosines lemmas revised for errata

# Main

Let's consider a $`r N_DIMS`$ latent space $`r LATENT_SPACE`$ with Euclidean
vector structure, and let $`r LS_BASIS_EQ`$ be a basis of $`r LATENT_SPACE`$.

@@

Let's consider a dichotomous item $`r ITEM_INDEX`$ such that the probability
that respondent with trait $`r LS_BASIS_EQ`$ gives a positive[^1] response is
given by

@@

Let's consider in $`r LATENT_SPACE`$ the standard basis $`r LS_STD_BASIS_EQ`$ as
well.

@@

$$
  `r BASIS_CHANGE_EQ`
$$ {#eq-change-basis}

@@

In $`r LATENT_SPACE`$ with the standard basis we have the usual scalar product.
In $`r LATENT_SPACE`$ with basis $`r LS_BASIS`$ we define an inner product
$`r INNER_PROD_BASIS`$ such that the norm of $`r TRAIT_VECTOR`$ is invariant,
i.e., $`r TRAIT_NORMS_EQ`$. Thus,[^4]

@@

Therefore we must define the inner product $`r INNER_PROD_BASIS`$ with Gram
matrix $`r INNER_PROD_MAT_EQ`$. That is, $`r INNER_PROD_EQ`$.[^5]

The Gram matrix has as element $`r INNER_PROD_MAT_EL_EQ`$. As
$`r TRANSFORM_MATRIX`$ (and consequently, $`r TRANSFORM_MATRIX_TRANSP`$) is
invertible so $`r INNER_PROD_MATRIX`$ is, and
$`r INNER_PROD_MATRIX_INV_EQ`$.[^6]

Let's consider now the vectors $`r DIR_COS_VEC`$ and $`r DIR_COS_VEC_STD`$, made
up by $`r N_DIMS`$ direction cosines of $`r TRAIT_VECTOR`$ with the basis
vectors of $`r LS_BASIS`$ and $`r LS_STD_BASIS`$, respectively.[^7]

::: {#lem-dircos}
Under the previously defined conditions, for all $`r TRAIT_VEC_IN_LS`$,[^8]

@@

::: {#lem-convert-spaces}
Let's consider another basis $`r NEW_BASIS_EQ`$ in $`r LATENT_SPACE`$ and the
inner product $`r INNER_PROD_NEWBS`$ that keeps the norm invariant.

Let $`r DIR_COS_VEC_NEWBS`$ be the vector of direction cosines of
$`r TRAIT_VECTOR`$ (i.e., the cosines with the elements of $`r NEW_BASIS`$).

@@

[^1]: I prefer to refer to the responses as "positive/negative", instead of
    "correct/incorrect", as this terminology contemplates the possibility of
    non-cognitive (as well as cognitive) domains.

[^2]: I'm preferring here the notation I use in the draft, that is,
    $`r DIM_INDEX`$ for indexing the dimensions (instead of $`r ITEM_INDEX`$,
    which is also used to index the items), $`r TRAIT_MODULE`$ for the
    coordinates of $`r TRAIT_VECTOR`$ in $`r LS_BASIS`$, and
    $`r TRAIT_VEC_STD_COMP`$ for the coordinates in $`r LS_STD_BASIS`$.
    Nevertheless, as Mario reminded me, it is worth noting that this notation
    for the vector coordinates is ambiguous, as we are using the same symbol
    ($`r TRAIT_VECTOR`$) for the "geometric vector" and the "algebraic vector"
    (i.e. column matrix) of its coordinates in $`r LATENT_SPACE`$ (this notation
    is usually accepted though).

[^3]: Is it ok to use square brackets here? I'm using it instead of parentheses
    because I had already written this (with a function), but if it's not
    correct I change it to parentheses.

@@

[^6]: I haven't considered it necessary to add the observation, as I think is a
    rather known theorem in matrix algebra (see e.g. [this response in Math
    Stackexchange](https://math.stackexchange.com/a/340234/1082326 "Response to 'Transpose of inverse vs inverse of transpose'")).

[^7]: I'm omitting here (for the moment) the mathematical expressions defining
    the direction cosine vectors.

[^8]: Mario is using $V$ here but I understand he refers to the latent (vector)
    space, which I have denoted it by $`r LATENT_SPACE`$. Is this correct?

>>>>

# Direction consines transformations

Let's consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$ with
Euclidean vector structure, and let $`r LS_BASIS_EQ`$ be a basis of
$`r LATENT_SPACE`$.

@@

Let's consider a dichotomous item $`r ITEM_INDEX`$ such that the probability
that a respondent with trait $`r TRAIT_VEC_IN_LS`$ gives a positive[^1] response
is given by

@@

Let's consider as well the standard basis $`r LS_STD_BASIS_EQ`$ in
$`r LATENT_SPACE`$.

@@

$$
  `r BASIS_CHANGE_EQ`.
$$ {#eq-change-basis}

@@

With the standard basis in $`r LATENT_SPACE`$ we have the usual scalar product.
With basis $`r LS_BASIS`$ in $`r LATENT_SPACE`$ we define an inner product
$`r INNER_PROD_BASIS`$ such that the norm of $`r TRAIT_VECTOR`$ is invariant,
i.e., $`r TRAIT_NORMS_EQ`$. Thus,[^4]

@@

Therefore we must define the inner product $`r INNER_PROD_BASIS`$ with Gram
matrix $`r INNER_PROD_MAT_EQ`$. That is, $`r INNER_PROD_EQ`$,[^5] which keeps
the norm invariant.

The Gram matrix has as element $`r INNER_PROD_MAT_EL_EQ`$.[^6] As
$`r TRANSFORM_MATRIX`$ (and consequently, $`r TRANSFORM_MATRIX_TRANSP`$) is
invertible so $`r INNER_PROD_MATRIX`$ is, and
$`r INNER_PROD_MATRIX_INV_EQ`$.[^7]

Let's consider now the vectors $`r DIR_COS_VEC`$ and $`r DIR_COS_VEC_STD`$, made
up by the $`r N_DIMS`$ direction cosines of $`r TRAIT_VECTOR`$ with the basis
vectors of $`r LS_BASIS`$ and $`r LS_STD_BASIS`$, respectively.[^8]

::: {#lem-dircos}
Under the previously defined conditions, for all $`r TRAIT_VEC_IN_LS`$,[^9]

@@

::: {#lem-convert-spaces}
Let's consider another basis $`r NEW_BASIS_EQ`$ in $`r LATENT_SPACE`$ and the
inner product $`r INNER_PROD_NEWBS`$ that keeps the norm of the Euclidean space
invariant.

Let $`r DIR_COS_VEC_NEWBS`$ be the vector of direction cosines of
$`r TRAIT_VECTOR`$ in $`r NEW_BASIS`$ (i.e., the cosines with the elements of
$`r NEW_BASIS`$).

@@

# Point of maximum slope

[^1]: I prefer to refer to the responses as "positive/negative", instead of
    "correct/incorrect", as this terminology allows for non-cognitive (as well
    as cognitive) domains.

[^2]: I'm preferring here the notation I use in the draft, that is,
    $`r DIM_INDEX`$ for indexing the dimensions (instead of $`r ITEM_INDEX`$,
    which is also used to index the items), $`r TRAIT_COMPONENT`$ for the
    coordinates of $`r TRAIT_VECTOR`$ in $`r LS_BASIS`$, and
    $`r TRAIT_VEC_STD_COMP`$ for the coordinates in $`r LS_STD_BASIS`$.
    Nevertheless, as Mario reminded me, it is worth noting that this notation
    for the vector coordinates is ambiguous, as we are using the same symbol
    ($`r TRAIT_VECTOR`$) for the "geometric vector" (i.e. "arrow" in vector
    space) and the "algebraic vector" (i.e. column matrix) of its coordinates in
    $`r LATENT_SPACE`$ (although this notation is usually accepted).

[^3]: Is it ok to use square brackets here? I'm using it instead of parentheses
    because I had already written this (with a function), but if it's not
    correct I change it to parentheses. **Also:** Is this a common algebra
    result? It feels like the this step is left implicit without justification;
    in my case, in the draft I needed to go all the way by Equations 2 through 4
    (section 2.2.1)

@@

[^6]: There is an erratum here in Mario's notes, I think. It should be the inner
    product of $`r BASIS_VECTOR_AUX_TRANSF`$ and $`r BASIS_VECTOR_ANY_TRANSF`$
    as I have written, instead of $`r BASIS_VECTOR_AUX`$ and
    $`r BASIS_VECTOR_ANY`$, as Mario wrote.

[^7]: I haven't considered it necessary to add the observation, as I think is a
    rather known theorem in matrix algebra (see e.g. [this response in Math
    Stackexchange](https://math.stackexchange.com/a/340234/1082326 "Response to 'Transpose of inverse vs inverse of transpose'")).

[^8]: I'm omitting here (for the moment) the mathematical expressions defining
    the direction cosine vectors.

[^9]: Mario is using $V$ here but I understand he refers to the latent (vector)
    space, which I have denoted it by $`r LATENT_SPACE`$. Is this correct?
